Liveness-Based Garbage Collection for Lazy Languages

### 2015-03-25 paper/paper.tex
Introducing 0bar 1 bar and 2bar

\newcommand{\scmin} {\mbox{\sf\em in}}
\newcommand{\scmnull}{\mbox{\sf\em null?}}
\newcommand{\scmpair}{\mbox{\sf\em pair?}}
\newcommand{\scmprim}{\ensuremath{\mathsf{+}}}
\newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}
\def\myvec{\mathaccent"017E } % wretched springer mode redefines \vec as bold!
\newcommand{\stk}{\mbox{S}}       % stack
\newcommand{\ID}{\mbox{$\mathbf{ id}$}} % identity function
\newcommand{\bang}{\mbox{\sc bang}}

\begin{document}
% --- Author Metadata here ---
\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%

\renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{0}\hspace*{-.5cm} (\DEFINE\ (\pf~\pa)
	  \UNL{1}\hspace*{-.5cm}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\CONS~1~\px) \IN
	  \UNL{2}\hspace*{-.5cm}      $\pi_2\!\!:\, $(\LET~\py\ $\leftarrow $\ (\CDR~\px) \IN  
          \UNL{3}\hspace*{-.5cm} \hspace*{.05cm}     $\pi_3\!\!:\,
          $(\LET\ \pz\  $\leftarrow$\   (\NULLQ~\pa) \IN
	  \UNL{4}\hspace*{-.45cm}      $\pi_4\!\!:\,
          $(\SIF\ $\pi_5\!\!:\,$ \pz\ ~$\pi_6\!\!:\,$ (\LET~\ww\  $\leftarrow$  (\take~5~\px) \IN
	  \UNL{5}\hspace*{.45cm}     $\pi_7\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\pw)  \IN
	  \UNL{6}\hspace*{.45cm}  $\pi_8\!\!:\,(\RETURN_1~\pu)$))
          \UNL{5}\hspace*{-.45cm}  $\pi_9\!\!:\, (\RETURN_2~\py)$)))))
	\end{uprogram}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\medskip

@@

Since   \pu\  is  $(\CDR~   \pw)$,  the   liveness  of   $\sigma$  for
\pu\ translates  into a demand  of $\sigma$ on $(\CDR~\pw)$,  which in
turn  generates  a  liveness  of $\{\pw.\epsilon,  ~\pw.1\sigma\}$  at
$\pi_7$.  The liveness of \pu\ should be killed at $\pi_7$ since it is
not in the scope of \pu.  However, this does not cause any imprecision
in the way  of the GC marking more  cells as live, since \pu\  is yet to
come into existence at $\pi_7$  and its liveness will not be consulted
by the garbage collector at this point.

@@

In an eager language, \pw\ would not have been live after $\pi_7$, the
point of  its last  use. In a  lazy language, however,  the definition
$\pu  \leftarrow (\CDR~\pw)$,  does  not result  in  an evaluation  of
$(\CDR~\pw)$; instead  a closure is created.   The evaluation actually
takes place while evaluating  $(\RETURN~\pu)$. Thus it is necessary to
record the liveness  of $\pw$ at the program  point $\pi_8$, indeed at
every program  point from $\pi_8$ up  to the beginning  of the program
(since  we do  not kill  liveness).  We  thus regard  the body  of the
function as  a tree and consider paths  from the root of  this tree to
program  points  which  trigger  a evaluation---\SIF\  conditions  and
\RETURN\  expressions.  We  call such  paths {\em  e-paths}.   For the
program shown, there are three such paths $EP_1 = \pi_1, \pi_2, \pi_3,
\pi_4,  \pi_5$, $EP_2  =  \pi_1, \pi_2,  \pi_3,  \pi_4, \pi_6,  \pi_7,
\pi_8$, and $EP_3 = \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.

Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\RETURN}.
The  expression $\pi:(\RETURN~x)$  with  a demand  $\sigma$ forces  an
evaluation  of  $x$  and  hence  a  new  map  ${\cal  M}=[\pi  \mapsto
  \{x.\sigma\}] $  needs to  be created.  The  ${\cal M}$-map  for the
expression $(\SIF~\pi:x~e_1~e_2)$  is a pointwise-union  of the ${\cal
  M}$-maps of $e_1$ and $e_2$.   In addition, since the condition also
triggers an evaluation of  $x$, the map $[\pi \mapsto \{x.\epsilon\}]$
is created and added to  the union.  Also notice that the consequences
of  laziness---the entire  expression including  the condition  is not
evaluated if  the demand  on it is  $\emptyset$.  This results  in the
empty map denoted by $\emptyset_{\cal{M}}$.

@@

Section~\ref{sec:computing}   shows   how   the  demand   transformers
$\Lfonly$  for  a  program  (representing  a  fully  context-sensitive
analysis) can  be safely  approximated, for each  function, by  a {\em
  procedure summary} (unifying the  contexts in the style of 0-CFA)\@.
The  summary consists  of a  pair  of a  single demand  and, for  this
demand, the corresponding  tuple of demands the function  makes on its
arguments.

%\comment{

\section{Generating Liveness Equations}
\begin{figure*}[t!]
\begin{picture}(100,130)(0,-90)
  \begin{tabular}{cc}
    \begin{boxedminipage}{.5\textwidth}
      {\sf
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\pi_3\!\!:\,$ \px
            \UNL{4} \hspace*{1.15cm} $\pi_4\!\!:\, (\RETURN_1~0)$
            \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\, (\RETURN_2~(1~+~\py))$)))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $\pi_\mainpgm\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
	  \UNL{2}   \hspace*{.78cm}              (\LET\ \pb\  $\leftarrow$  \ldots \IN 
          \UNL{3}   \hspace*{.78cm}    $\pi_8\!\!:\, $(\LET\ \pw\  $\leftarrow$  (\length\ \py) \IN~$\pi_{9}\!\!:\,$ (\RETURN\ \pw)))))))
\end{uprogram}}
        } 
    \end{boxedminipage}
    & 
    \begin{boxedminipage}{.5\textwidth}
      {\sf
        \begin{tabular}{lc}
        $\cal{M} = $ 
            \begin{tabular}{lc}            
              $[$
              $\pi_{\mbox{7}} \mapsto \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \
              \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace ,$ \\ 
              $\pi_{\mbox{4}} \mapsto \lbrace \rbrace , $\\
              $\pi_{\mbox{3}} \mapsto \lbrace \px.\epsilon, \xl.\epsilon \rbrace $  \\
              $]$
            \end{tabular} \\
            $\cal{P} = [$
              \begin{tabular}{lc}
              $\pi_{\mbox{1}} \mapsto \lbrace \pi_{\mbox{3}},\pi_{\mbox{4}}, \pi_{\mbox{7}} \rbrace, $ \
              $ \pi_{\mbox{2}} \mapsto \lbrace \pi_{\mbox{3}},\pi_{\mbox{4}}, \pi_{\mbox{7}} \rbrace, $ \\
              $ \pi_{\mbox{3}} \mapsto \lbrace \pi_{\mbox{3}} \rbrace, $ \
              $ \pi_{\mbox{4}} \mapsto \lbrace \pi_{\mbox{4}} \rbrace, $ \
              $ \pi_{\mbox{5}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace, $ \\
              $ \pi_{\mbox{6}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace, $ \
              $ \pi_{\mbox{7}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace $ \\
              $]$
            \end{tabular}
        \end{tabular}
      }
    \end{boxedminipage}
    \\
    (a) Example program.&
    \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
        (b) {\em e-path} to liveness environment map \\
        (c) Prog pt to {\em e-paths } map
    \end{tabular}}
  \end{tabular}
\end{picture}
\kern -3ex

>>>>

%\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
\newcommand{\scmin} {\mbox{\sf\em in}}
\newcommand{\scmnull}{\mbox{\sf\em null?}}
\newcommand{\scmpair}{\mbox{\sf\em pair?}}
\newcommand{\scmprim}{\ensuremath{\mathsf{+}}}
\newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}
\def\myvec{\mathaccent"017E } % wretched springer mode redefines \vec as bold!
\newcommand{\stk}{\mbox{S}}       % stack
\newcommand{\ID}{\mbox{$\mathbf{ id}$}} % identity function
\newcommand{\bang}{\mbox{\sc bang}}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition} 
\begin{document}
% --- Author Metadata here ---
\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[t!]
\begin{picture}(100,130)(0,-90)
  \begin{tabular}{cc}
    \begin{boxedminipage}{.5\textwidth}
      {\sf
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} $(\DEFINE\ (\length~\xl)$
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow \ (\NULLQ~\xl)~\IN$
            \UNL{3} \hspace*{.05cm} $\psi_1\!\!:\,$(\SIF\ $\pi_3\!\!:\,$ \px
            \UNL{4} \hspace*{1.15cm} $\pi_4\!\!:\, (\RETURN_1~0)$
            \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\py\  $\leftarrow  (\CDR~\xl)$  \IN
            \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\,
            $(\LET~\pz\  $\leftarrow  (\length~\py)~\IN$
            \UNL{6} \hspace*{1.15cm}  $\pi_7\!\!:\,
           $(\LET~\pw\  $\leftarrow  (\PRIM~1~\pz)~\IN$
            \UNL{7} \hspace*{.05cm}\hspace*{1.15cm} $\psi_2\!\!:\, (\RETURN_1~\pw)))))))$
        \end{uprogram}}

        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $\pi_\mainpgm\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
	  \UNL{2}   \hspace*{.78cm}              (\LET\ \pb\  $\leftarrow$  \ldots \IN 
          \UNL{3}   \hspace*{.78cm}    $\pi_8\!\!:\,
          $(\LET\ \pc\  $\leftarrow (\length\ \pb)~\IN$
          \UNL{4}\hspace*{.05cm}\hspace*{.78cm} $\psi_{3}\!\!:\, (\RETURN\ \pc)))))))$
\end{uprogram}}
        } 
    \end{boxedminipage}
    & 
    \begin{boxedminipage}{.5\textwidth}
      {\sf
        \begin{tabular}{lc}
        $\cal{M} = $ 
            \begin{tabular}{lc}            
              $[$
              $\pi_{\mbox{7}} \mapsto \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \
              \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace ,$ \\ 
              $\pi_{\mbox{4}} \mapsto \lbrace \rbrace , $\\
              $\pi_{\mbox{3}} \mapsto \lbrace \px.\epsilon, \xl.\epsilon \rbrace $  \\
              $]$
            \end{tabular} \\
            $\cal{P} = [$
              \begin{tabular}{lc}
              $\pi_{\mbox{1}} \mapsto \lbrace \pi_{\mbox{3}},\pi_{\mbox{4}}, \pi_{\mbox{7}} \rbrace, $ \
              $ \pi_{\mbox{2}} \mapsto \lbrace \pi_{\mbox{3}},\pi_{\mbox{4}}, \pi_{\mbox{7}} \rbrace, $ \\
              $ \pi_{\mbox{3}} \mapsto \lbrace \pi_{\mbox{3}} \rbrace, $ \
              $ \pi_{\mbox{4}} \mapsto \lbrace \pi_{\mbox{4}} \rbrace, $ \
              $ \pi_{\mbox{5}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace, $ \\
              $ \pi_{\mbox{6}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace, $ \
              $ \pi_{\mbox{7}} \mapsto \lbrace \pi_{\mbox{7}} \rbrace $ \\
              $]$
            \end{tabular}
        \end{tabular}
      }
    \end{boxedminipage}
    \\
    (a) Example program.&
    \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
        (b) {\em e-path} to liveness environment map \\
        (c) Prog pt to {\em e-paths } map
    \end{tabular}}
  \end{tabular}
\end{picture}
\kern -3ex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%% %%%%%%%%%%%%%%%%%%%%%%%%

%% \renewcommand{\arraystretch}{1}{
%% 	  \begin{uprogram}
%% 	  \UNL{0}\hspace*{-.5cm} (\DEFINE\ (\pf~\pa)
%% 	  \UNL{1}\hspace*{-.5cm}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\CONS~1~\px) \IN
%% 	  \UNL{2}\hspace*{-.5cm}      $\pi_2\!\!:\, $(\LET~\py\ $\leftarrow $\ (\CDR~\px) \IN  
%%           \UNL{3}\hspace*{-.5cm} \hspace*{.05cm}     $\pi_3\!\!:\,
%%           $(\LET\ \pz\  $\leftarrow$\   (\NULLQ~\pa) \IN
%% 	  \UNL{4}\hspace*{-.45cm}      $\pi_4\!\!:\,
%%           $(\SIF\ $\pi_5\!\!:\,$ \pz\ ~$\pi_6\!\!:\,$ (\LET~\ww\  $\leftarrow$  (\take~5~\px) \IN
%% 	  \UNL{5}\hspace*{.45cm}     $\pi_7\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\pw)  \IN
%% 	  \UNL{6}\hspace*{.45cm}  $\pi_8\!\!:\,(\RETURN_1~\pu)$))
%%           \UNL{5}\hspace*{-.45cm}  $\pi_9\!\!:\, (\RETURN_2~\py)$)))))
%% 	\end{uprogram}}
%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% \medskip

@@

Since   \pu\  is  $(\CDR~   \pw)$,  the   liveness  of   $\sigma$  for
\pu\ translates  into a demand  of $\sigma$ on $(\CDR~\pw)$,  which in
turn  generates  a  liveness  of $\{\pw.\epsilon,  ~\pw.1\sigma\}$  at
$\pi_7$.  The liveness of \pu\ should be killed at $\pi_7$ since it is
not in the scope of \pu.  However, this does not cause any imprecision
in the way  of the GC marking more  cells as live, since \pu\  is yet to
come into existence at $\pi_7$  and its liveness will not be consulted
by the garbage collector at this point.

In an eager language, \pw\ would not have been live after $\pi_7$, the
point of  its last  use. In a  lazy language, however,  the definition
$\pu  \leftarrow (\CDR~\pw)$,  does  not result  in  an evaluation  of
$(\CDR~\pw)$; instead  a closure is created.   The evaluation actually
takes place while evaluating  $(\RETURN~\pu)$. Thus it is necessary to
record the liveness  of $\pw$ at the program  point $\pi_8$, indeed at
every program  point from $\pi_8$ up  to the beginning  of the program
(since  we do  not kill  liveness).  We  thus regard  the body  of the
function as  a tree and consider paths  from the root of  this tree to
program  points  which  trigger  a evaluation---\SIF\  conditions  and
\RETURN\  expressions.  We  call such  paths {\em  e-paths}.   For the
program shown, there are three such paths $EP_1 = \pi_1, \pi_2, \pi_3,
\pi_4,  \pi_5$, $EP_2  =  \pi_1, \pi_2,  \pi_3,  \pi_4, \pi_6,  \pi_7,
\pi_8$, and $EP_3 = \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.

Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\RETURN}.
The  expression $\pi:(\RETURN~x)$  with  a demand  $\sigma$ forces  an
evaluation  of  $x$  and  hence  a  new  map  ${\cal  M}=[\pi  \mapsto
  \{x.\sigma\}] $  needs to  be created.  The  ${\cal M}$-map  for the
expression $(\SIF~\pi:x~e_1~e_2)$  is a pointwise-union  of the ${\cal
  M}$-maps of $e_1$ and $e_2$.   In addition, since the condition also
triggers an evaluation of  $x$, the map $[\pi \mapsto \{x.\epsilon\}]$
is created and added to  the union.  Also notice that the consequences
of  laziness---the entire  expression including  the condition  is not
evaluated if  the demand  on it is  $\emptyset$.  This results  in the
empty map denoted by $\emptyset_{\cal{M}}$.

@@

@@

Section~\ref{sec:computing}   shows   how   the  demand   transformers
$\Lfonly$  for  a  program  (representing  a  fully  context-sensitive
analysis) can  be safely  approximated, for each  function, by  a {\em
  procedure summary} (unifying the  contexts in the style of 0-CFA)\@.
The  summary consists  of a  pair  of a  single demand  and, for  this
demand, the corresponding  tuple of demands the function  makes on its
arguments.
\section{Computing liveness and its encoding as a table}\label{sec:computing}

Section~\ref{sec:liveness} gave a context-sensitive liveness
analysis and {\color {red}  proved it correct with reference
  to   a  {\em   minefield}   semantics}.   In   particular,
\mmap{f}{\sigma}{\Lfonly}  (and  the  map \pmap{f}  together
describes  a  liveness  set  for  each program  point  in  a
function in terms of a  given $\sigma$ and \Lfonly. We still
have not mentioned what is the specific sigma that should be
used for each function How is LF to be computed.

It helps to cast the equations  in the form of a grammar. To
do  this  we  first  have  to  cast  massage  the  equations
slightly:

\subsubsection{Modifying the liveness rules.}

The      $\mathit{ref}$     rule      for     \CONS,      shown     in
Figure~\ref{fig:live-judge},  requires   us  to  remove   the  leading
\acar\ and \acdr\ from  access paths in $\sigma$.  Mathematically this
is fine but  causes problems when solving the  liveness equations {\em
  symbolically}  since such  decomposition  cannot be  expressed as  a
context-free  grammar.  Similarly,  the rules  for \CAR,  \CDR, \PRIM,
\NULLQ,  and \SIF\  require  us  to return  $\emptyset$,  on the  null
demand. While this is mathematically fine, for our grammar formulation
we would like to treat these operations symbolically.

To  handle  this,  we  first  introduce  three  new  symbols.
First, \bcar\ and \bcdr\ are defined as:
\begin{align*}
 \bcar\sigma &\triangleq \{\alpha \mid 0\alpha \in \sigma\}\\
 \bcdr\sigma &\triangleq \{\alpha \mid 1\alpha \in \sigma\}
\end{align*}

and \cbar\ as:

\begin{align*}
 \cbar\sigma \triangleq& \emptyset~\mbox{if}~\sigma = \emptyset\\
                       & \{\epsilon\} ~\mbox{otherwise}
\end{align*}

We can now rewrite the \CONS\ and the \CAR\ rules of $\mathit{ref}$ as:
\begin{align*}
&\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
= x.\bcar\sigma \cup y.\bcdr\sigma  \label{eqn:mod-cons},~
\mbox{and} \\
&\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          =   x.\cbar\sigma \cup x.0\sigma
\end{align*}
and the \Lfunonly\ rule
for \SIF\ as:
\begin{align*}
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) =
                    &\mathcal{L}(e_1,\sigma,\Lfonly)~\uplus
        \mathcal{L}(e_2,\sigma,\Lfonly)~\uplus\\
        &[\,l \mapsto  x.\cbar \sigma]
\end{align*}

The  defintions of \bcar\  and  \bcdr\  induce   the  following
relation $\hookrightarrow$ over sets of access paths:
\begin{align*}
  \sigma_1\bcar\sigma_2  &\hookrightarrow 
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \bcar\sigma_2',\\
\sigma_1\bcdr\sigma_2  &\hookrightarrow 
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
  \bcdr\sigma_2',~\mbox{and}\\  
\sigma_1\cbar \emptyset &\hookrightarrow \emptyset
\end{align*}

We  call the liveness equations with  this modification
  \Lfunonly$'$.

The reflexive  transitive closure of  $\hookrightarrow$ will
be   denoted    as   $\stackrel{*}{\hookrightarrow}$.    The
following proposition relates  the original and the modified
rules.

\begin{proposition}
Assume that a liveness computation based on the original set
of rules gives the liveness of the variable $x$ at a program
point  $\pi_i$   as  $\sigma$  (symbolically,  $\Lanv{i}{x}=
\sigma$).   Further,  let  $\Lanv{i}{x}= \sigma'$  when  the
modified rules are used instead of \Lfunonly.  Then $\sigma'
\stackrel{*}{\hookrightarrow} \sigma$.
\end{proposition}

%% To see why the  proposition is true, consider an analysis
%% based on  \Lfunonly$'$ in  which $\sigma$ appears  in the
%% context      $\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)$.
%% Let $\alpha \in  \sigma$.  The symbol \bcar(\bcdr) merely
%% marks a place in  $\alpha$ where the original \CONS\ rule
%% would have erased  an immediately following \acar(\acdr),
%% or,  in absence  of  such a  symbol,  would have  dropped
%% $\alpha$ itself.   Since the  application of any  rule in
%% \Lfunonly$'$  merely  adds symbols  at  the beginning  of
%% $\alpha$, the  markers and other symbols  in $\alpha$ are
%% propagated to  other dependent parts of  program in their
%% same  relative   positions.   Consequently,  the  erasure
%% carried   out   at  the   end   of   the  analysis   with
%% $\stackrel{*}{\hookrightarrow}$ gives  the same result as
%% obtained through \Lfunonly.

\alan{For practical use we need to solve the liveness equations {\em finitely and
symbolically}.  As expressed mathematically, and given a fixed program,
three things are potentially unbounded:
({\em i\/}) the number of call strings (and hence arguments $\sigma$ to $\Lfonly$);
({\em ii\/}) the length of access paths $\alpha \in \sigma$ and
({\em iii\/}) the number of such access paths.

We commonly first solve ({\em i\/}) by reducing the number of distinct
calling contexts (e.g.\  to a single unified context  in 0-CFA style).
However, it  turns out that we  can solve the  equations for $\Lfonly$
symbolically  without  this  reduction,  so  here  we  defer  this  to
Section~\ref{sec:bodylivenessbodies}.    (It    also   allows   easier
extension to dynamically determined liveness---future work.)
%
We address ({\em ii\/}) and ({\em iii\/}) by 
% attempting to re-interpret
re-interpreting
the liveness definitions in Figure~\ref{fig:live-judge} symbolically as a grammar
rather than a mutually recursive set of equations on sets of access paths.

This requires two ideas.  Firstly we have to control the use of functions---they
tend to be infinitary and do not occur naturally in CFGs; in particular our
$\Lv$ maps names to access paths, and  $\Lfonly$ maps access paths
to a tuple of access paths.  The former is achieved by using a separate
meta-variable (later non-terminal)
$\Lanv{i}{x}$ for each variable $x$ and each program point $\pi_i$
($\Lanv{i}{x}$ represents $\Lv(x)$ at $\pi_i$).
Section~\ref{sec:Lfonlysymbolically} shows how the latter $\Lfonly_f^i$ is also
expressible finitely (it is a linear form).

Secondly, there are also two technical issues in re-interpreting
Figure~\ref{fig:live-judge} as a grammar.
One is the use of ``$\cdots \setminus \{x.*\}$'', but this restricted form of `$\setminus$'
disappears after variables are separated ($\Lanv{i}{x}$ above). The other
is that $\mathit{ref}$ for $\CONS$ decomposes strings and thus gives a general grammar
not a CFG\@.
Below we show how symbols $\bcar$, $\bcdr$ can give an equivalent CFG\@.

Finally,  Section~\ref{sec:NFA-approx}
uses a construction due to
Mohri and Nederhof~\cite{mohri00regular} to over-approximate
context-free grammars with regular grammars;
these are more appropriate for run-time use.
%
Hence the overall `big picture' view is that each GC point
is annotated with a table encoding the DFA for that program point.
Such program points forming the call string causing the garbage collection
(and ending in $\CONS$)
are used by the garbage collector to curtail its reachability-marking phase
to this over-approximated liveness.

\subsubsection{Modifying the \CONS\ rule.}

The    $\mathit{ref}$    rule for \CONS,    shown    in
Figure~\ref{fig:live-judge},  requires  us  to remove  the  leading
\acar\  and  \acdr\  from  access paths in  $\sigma$.
  Mathematically  this  is fine  but causes problems  when
  solving  the liveness equations  {\em symbolically}  since such
  decomposition cannot be expressed as a context-free grammar.  
To  handle this,  we introduce  two  new symbols  \bcar\  and
\bcdr\ with the properties:
\[
 \bcar\sigma \triangleq \{\alpha \mid 0\alpha \in \sigma\}
\hspace{3em} \mbox{and} \hspace{3em}
 \bcdr\sigma \triangleq \{\alpha \mid 1\alpha \in \sigma\}
\]
We can now rewrite the \CONS\ rule as:
\begin{eqnarray*}
\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
&=& x.\bcar\sigma \cup y.\bcdr\sigma  \label{eqn:mod-cons}
\end{eqnarray*}

We  call the liveness equations with  this modification
  \Lfunonly$'$.
The  defintions of \bcar\  and  \bcdr\  induce   the  following
relation $\hookrightarrow$ over sets of access paths:
\begin{eqnarray*}
  \sigma_1\bcar\sigma_2  \hookrightarrow 
  \sigma_1\sigma_2'\mbox{ where } \sigma_2' = \bcar\sigma_2
& \qquad\mbox{and}\qquad &
\sigma_1\bcdr\sigma_2  \hookrightarrow 
  \sigma_1\sigma_2'\mbox{ where } \sigma_2' = \bcdr\sigma_2  
\end{eqnarray*}

The reflexive  transitive closure of  $\hookrightarrow$ will
be   denoted    as   $\stackrel{*}{\hookrightarrow}$.    The
following proposition relates \Lfunonly\ and \Lfunonly$'$:

\begin{proposition}
Assume that a liveness computation based on \Lfunonly\ gives
the liveness of the variable  $x$ at a program point $\pi_i$
as $\sigma$ (symbolically, $\Lanv{i}{x}= \sigma$).  Further,
let $\Lanv{i}{x}= \sigma'$ when \Lfunonly$'$ is used instead
of  \Lfunonly.  Then  $\sigma' \stackrel{*}{\hookrightarrow}
\sigma$.
\end{proposition}

To  see why the  proposition is  true, consider  an analysis
based  on  \Lfunonly$'$ in  which  $\sigma$  appears in  the
context  $\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)$.   Let
$\alpha \in \sigma$.  The symbol \bcar(\bcdr) merely marks a
place in $\alpha$ where  the original \CONS\ rule would have
erased an immediately following \acar(\acdr), or, in absence
of such a symbol, would have dropped $\alpha$ itself.  Since
the  application of  any  rule in  \Lfunonly$'$ merely  adds
symbols at the beginning  of $\alpha$, the markers and other
symbols in $\alpha$ are  propagated to other dependent parts
of program in  their same relative positions.  Consequently,
the  erasure carried  out at  the end  of the  analysis with
$\stackrel{*}{\hookrightarrow}$  gives  the  same result  as
obtained through \Lfunonly.

\subsection{Generating equations for the demand transformer \Lfonly}
\label{sec:Lfonlysymbolically}

%\marginpar{Nearly same as Figure~\ref{fig:mot-example}!}
We shall consider  the program in Figure~\ref{fig:examp-pgm}
as a running example; this is the ANF version of the program
in Figure~\ref{fig:mot-example}.

%%AM no longer used: In  what follows, we shall overload the semantics
%% of   \Lfunonly\  so  that   \Lfun{e}{\sigma}{\Lfonly}  additionally
%% records  the liveness  environment at  every sub-expression  of $e$
%% during its computation.
%mean the generation
%of equations $\Lanv{i}{x}  = \ldots$, where $\pi_i$ is  the label of a
%sub-expression of $e$ and $x$ is a variable in scope at $\pi_i$.

To generate the equations defining \Lfonly$_{\mathit  f}$, we follow the rule  \rn{define-live}.   We start with a {\em symbolic demand}
$\sigma$  and determine $\Lv = \Lfun{e_{\mathit{f}}}  {\sigma}
{\Lfonly}$, treating \Lfonly\  as an uninterpreted function symbol.  We  then generate  equations of  the form
$\Lf{\mathit{f}}{i}{\sigma} = \Lv(x_i)$  where $x_i$ is the $i$th
formal parameter  of $\mathit{f}$ and $\Lv(x_i)$  is the liveness
of $x_i$.  For our example program which has  a
single function \append, this generates the following equations:
% $\Lfonly_{\append}$  is defined by equations:
  \begin{eqnarray*}
    \Lf{\append}{1}{\sigma}
    &=& \epsilonset \cup \acar\bcar\sigma
    \cup \acdr\Lf{\append}{1}{\bcdr\sigma}\\ 
    \Lf{\append}{2}{\sigma}
    &=& \sigma \cup \Lf{\append}{2}{\bcdr\sigma} 
  \end{eqnarray*}
  In  general, the  equations  for \Lfonly\  are recursive  since
  \Lv\ may, in turn, be expressed in terms of \Lfonly.  We assume
  that $\Lfonly_{\mathit f}$ is expressible in the closed form as:\footnote{This
    is  similar to  solving the  differential equation  
    $\mathit{ay}''+\mathit{by}'+\mathit{c}  = 0$, where  we guess
    that    the    solution     has    the    form    $\mathit{y}
    =\mathit{e}^{\mathit{rx}}$.  Substituting the solution in the
    equation  yields  a  quadratic  equation  in  $r$,  and  each
    solution of $r$ gives rise  to a solution of the differential
    equation  (in our  setup we  can effectively  pick  the least
    solution rather than needing linear combinations).  }
\begin{eqnarray}
\label{eq:LF:DI}
  \Lf{f}{i}{\sigma} &=& \Uf{f}{i} \cup \Df{f}{i}.\sigma
\end{eqnarray}
where  \Uf{f}{i}  and \Df{f}{i}  are  sets  of  strings over  the
alphabet $\lbrace \acar, \acdr,\bcar, \bcdr \rbrace$.  The reason
why \Lfonly\ has this form  is as follows: Recall that
\Lfonly$_{\mathit  f}$ describes  the demand  transformation from
the call of a function ${\mathit f}$ to its arguments. \Uf{f}{i} represents a
situation in which  the demand $\sigma$ does not  reach the $i$th
argument.   This happens,  for  instance, when  argument is  used
only within  the condition  of  an \SIF.   $\Df{f}{i}$ represents  the
transformation of the demand $\sigma$ before it reaches the $i$th
argument. This  transformation prefixes strings in 
$\sigma$ by a string from $\{\acar,\acdr,\bcar,\bcdr\}^*$.

Substituting  the  guessed   form  in  the  equations  describing
\Lfonly$_{\mathit   f}$,  and   separately  equating   the  terms
containing $\sigma$ and those  without $\sigma$ on both sides of the
resulting equation, we
get equations for \Uf{f}{i} and \Df{f}{i} that are independent of
% $\sigma$.   Any  pair of solutions to  \Uf{f}{i}  and  \Df{f}{i} yields  a
$\sigma$.  Any solution for  \Uf{f}{i}  and  \Df{f}{i} yields  a
solution for  \Lfonly$_{\mathit f}$. Applied  to \Lfonly$_{\mathit
  \append}$, we get:
  \begin{align*}
    \Uf{\append}{1} &= \epsilonset
    \cup \acdr.\Uf{\append}{1}  
    &\quad 
    \Uf{\append}{2} &= \Uf{\append}{2} \\
    \Df{\append}{1} &= \lbrace\acar\bcar\rbrace 
    \cup \acdr\Df{\append}{1}\bcdr &\quad 
    \Df{\append}{2} &= \epsilonset \cup
    \Df{\append}{2}\bcdr 
  \end{align*}
Note that these can be viewed as CFGs, with all but $\Df{\append}{1}$ being regular.

\subsection{Generating liveness equations \Lv\  for function bodies}
\label{sec:bodylivenessbodies}
